Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-06-16 paper/paper.tex
I've made one pass

\section{Recursion Schemes}\label{sec:category-theory}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes, which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
further these ideas to account for generic (co)inductive data types and generic
algorithms operating on them.

@@

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by $\Arr
(\CatC)$, such that there always exists an identity arrow $id_{A} : A \to A$ for
each object $A$ and for two arrows $A \xrightarrow{f} B$ and $B \xrightarrow{g}
C$ there always exists an arrow $A \xrightarrow{g\circ f} C$ obeying the
associativity law. We denote $\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to
$B$ and we use the letters $\CatC$, $\CatD$, $\CatE\dots$ for
categories\footnote[1]{For presentation purposes we shall not deal with size
  issues and assume all the categories are locally small.}. The initial object
in a category, denoted by $0$, is the object such that for any other object $A$
there is a unique arrow $0 \xrightarrow{!} A$.  Dually, the terminal object,
denoted by $1$, is the object such that for any other object $A$ there is a
unique arrow $A \xrightarrow{!} 1$. As a result of the uniqueness properties, initial and terminal objects are unique up-to
isomorphism.

@@

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$. Intuitively, a natural
transformation transforms the structure of a functor into the structure of
another functor without assuming what is the type of data contained in them, for
example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
nothing if the list is empty and the head of the list otherwise needs not to
know the type of the data.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For a functor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}. These are  arrows $f : X \to Y$ in the underlying category $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = a_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
if $F$ has an initial $F$-algebra then this is the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ where
$\InIso : F \mu F \to \mu F$ is the $F$-algebra  witnessing the
isomorphism $F\mu F \iso \mu F$, furthermore  $\InOp : \mu F \to F \mu F$ is the
inverse of $\InIso$. The uniqueness property of initial $F$-algebras states that for any other
$F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra homomorphism, denoted
by $\cata{\beta} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
``\emph{catamorphism}'' or ``\emph{fold}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InOp
\end{equation}
As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see.

@@

For example, for a set of states $X$ and a finite set of labels $L$ we can
define a labelled transition system (LTS) on $X$ as a function $X \to BX$
implementing the \emph{transition system} with $BX = L \times X$.  In
particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.
In this case, we can even set some notation for the transition map:
\[
  x \xrightarrow{l} y \eqdef c_{X}(x) = (l, y)
\]
The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
are $B$-coalgebras and morphisms $f : (X, c_{X}) \to (Y, c_{Y})$ are
$B$-coalgebra homomorphisms $f : A \to B$, that is
$ c_{Y} \circ f = F(f) \circ c_{X}$. The terminal object in this category is
called the terminal, or final, $B$-coalgebra. The carrier of this coalgebra
corresponds to the greatest fixed-point for the functor $B$, denoted by
$(\nu B, \OutIso)$ with $\OutIso$ being the final $B$-coalgebra witnessing the
isomorphism and $\OutIso^{\circ} : B\nu B \to \nu B$ being its inverse.

For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
of infinite streams over the set $A$, that is the greatest solution to the domain equation
$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
a unique $B$-coalgebra homomorphism into the terminal coalgebra
$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
\end{equation}
From the uniqueness property we can derive the fusion law for unfolds:
\begin{equation}
  \label{eq:ana:fusion}
  \ana{c_{Y}} \circ f = \ana{c_{X}}
\end{equation}
for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.

@@

A coalgebra $(C,c)$ is \emph{recursive} if for every algebra $(A, a)$ there is a
\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$. We denote these type of
hylos by $\hylo{c}{a}$.

@@

Nevertheless, since \coq{partition} always produces sublists that are strictly
smaller than the input list, it still supports a well-founded recursion scheme,
ensuring the existence of a unique solution. The uniqueness property of the
hylomorphisms yields the following fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
optimisation}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. This, in turn,
allows us to prove that a recursive hylomorphism is the composition of a
catamorphism and a recursive anamorphism.

@@

\subsection{Polynomial Functors}
Recursion schemes are formulated using generic functors that encapsulate the
structure of the recursive operator. However, since not all functors have the
suitable fixed points, it is necessary to restrict our focus to \emph{strictly
positive functors}.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation of hylomorphisms in Coq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}), then we formalise coalgebras
(Section~\ref{sec:coq-coalg}) and then we put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

@@

Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*...*)
\end{coqcode}
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that
is an \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same
coalgebra.

@@

% \subsection{Recursive Anamorphisms}
% Anamorphisms applied to recursive coalgebras specialise to hylomorphisms into an
% inductive data type in the following way.  Intuitively, a recursive coalgebra
% can be applied only finitely many times, therefore when this is applied to an
% anamorphism the only possible traces it can produce from the seed function are
% the finite ones. We denote this special kind of recursion scheme \emph{recursive
% anamorphism}.  We can show that recursive anamorphisms of type \coq{X -> GFix F}
% can also be given the type \coq{X -> LFix F}. Moreover, these anamorphisms are
% exactly hylomorphisms on the recursive \coq{F}-coalgebra and the
% algebra \coq{In} for the inductive data type \coq{LFix F}. This fact falls out
% from the universality property of the hylomorphism and the fact that recursive
% anamorphisms satisfy the same equation.
%
% We now show how this is proven in the Coq development. First we define a
% recursive anamorphism as a function into \coq{LFix F}:
% \begin{coqcode}
% Definition rana_f__ `{setoid A} (c : Coalg F A)
%   : forall x : A, RecF c x -> LFix F
%   := fix f (x : A) (H : RecF c x) :=
%        let hx := c x in
%        LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
%
% Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F := ...
% \end{coqcode}
% We omit the boilerplate details to obtain the setoid morphism.
% The recursive anamorphism has the following universality property:
% \begin{coqcode}
% Lemma rana_univ A {eA : setoid A} (h : RCoalg A) (f : A ~> LFix F)
%     : f =e rana h <-> f =e l_in \o fmap f \o h.
% \end{coqcode}
% We can now show that every hylomorphism is composed by a recursive anamorphism
% followed by a catamorphism:
% \begin{coqcode}
% Corollary cata_ana_hylo `(F : Cont Sh P) `{setoid A} `{setoid B}
%   (g : Alg F B) (h : RCoalg F A)
% : cata g \o rana h =e hylo g h.
% \end{coqcode}
% This result will allow us to \emph{calculate} programs by providing small, easy
% to verify implementations, and combining them into efficient recursive functions
% by using Coq's proofs of hylomorphism laws as optimisation passes.

\subsubsection{On the subtype of finite elements}
As we mention in
Section~\ref{sec:category-theory}, we define recursive anamorphisms as
hylomorphisms built by using a recursive coalgebra, and the respective
initial \coq{F}-algebra.
In other words, in this development we have defined recursive anamorphisms on
inductive data types.  We might have as well defined them on the subtype of
finite elements of coinductive data types using a predicate which states when an
element of a coinductive data type is finite:
\begin{coqcode}
Definition FinF : GFix F -> Prop := RecF g_out.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}.
We show this by
defining a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as hylomorphisms on the
recursive recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
\begin{coqcode}
Definition rana c := hylo l_in c.
\end{coqcode}
The main advantage of this definition is that recursive anamorphisms compose
with catamorphisms without the need to reason about termination and finiteness
of values.
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
\coq{times_two} multiplies every element of the list by two. We can use Coq's
generalised rewriting and \coq{hylo_fusion_l} to fuse \coq{times_two} into
\coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by applying
\coq{times_two} to the shapes thus multiplying every pivot by two.

@@

In our framework, we formalise $G$ as a container with shape \coq{Sg} and
position \coq{Pg}, then \coq{Memo} is the least fixed-point of the functor
$F X = A \times GX$:
\begin{coqcode}
Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table := LFix Memo.
\end{coqcode}
The operations of the cofree comonad \coq{Memo} are the \coq{Cons}, the \coq{headT} and the \coq{tailT}:
\begin{coqcode}
Definition consT : A * App G Table ~> Table := (* *)
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
Finally, we define a dynamorphism as a hylomorphism from a type \coq{B} into a
type \coq{Table} post-composed with the \coq{headT} map which outputs the last
result from the memory table:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (consT \o pair a id) c.
\end{coqcode}
Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$
is not a $G$-algebra.  To recover the $G$-algebra we pair it with the identity
map and  so we need to transmogrify it by pairing it with the identity map and
post-composing it with \coq{consT}. This algebra uses this table to lookup
elements, instead of triggering a further recursive call, then elements are
inserted into the memoisation table by the use of \coq{consT} to the result of
applying the algebra. In our example this algebra is called \coq{knapsackA} (see
accompanying code).
% \begin{coqcode}
% Fixpoint memo_knap table wvs :=
%   match wvs with | nil => nil | h :: t =>
%       match lookupT (Nat.pred (fst h)) table with
%       | Some u => (u + snd h)%sint63 :: memo_knapsack table t
%       | None => memo_knapsack table t
%       end
%   end.
% Definition knapsack_alg (wvs : list (nat * int))
%   (x : App NatF (Table NatF int)) : int :=
%   match x with | MkCont sx kx => match sx with
%   | inl tt => fun kx => 0%sint63
%   | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
%   end kx end.
% Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
% \end{coqcode}
Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we
can define knapsack using the recursion scheme for dynamorphisms:
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
The full extracted code coming from this specification is available in the
artefact accompanying this paper, and an interested reader can check that simple
inlining would produce the following:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note that in the extracted code, the recursive calls to \ocaml{f} build the
memoisation table, and that this memoisation table is used to compute the
intermediate results in \ocaml{memo_knap}, which is finally discarded to produce
the final result.

>>>>

\section{Recursion Schemes}\label{sec:category-theory}
The structure of data is very similar to the structure of an algorithm which
processes that data. This relationship manifests in the form of structured
recursion schemes, which are widely used in functional languages such Haskell.
Canonical examples are folds (catamorphisms), which consume data, and unfolds
(anamorphisms), which produce it. While some implementations of recursion
schemes like $\operator{foldr}$/$\operator{unfoldr}$ in Haskell are specific to
a particular data structure, in this case $\operator{List}$s, we can generalise
these ideas further to account for generic (co)inductive data types and generic
algorithms operating on them.

@@

\subsection{Elements of Category Theory}
A \emph{category} is a collection of objects $A, B, C$, denoted by $\Obj(\CatC)$
and a collection of arrows $f, g, h$ between these objects, denoted by $\Arr
(\CatC)$, such that there always exists an identity arrow $id_{A} : A \to A$ for
each object $A$ and for two arrows $A \xrightarrow{f} B$ and $B \xrightarrow{g}
C$ there always exists an arrow $A \xrightarrow{g\circ f} C$ obeying the
associativity law. We denote $\Hom_{\CatC} (A, B)$ the set of arrows from $A$ to
$B$ and we use the letters $\CatC$, $\CatD$, $\CatE\dots$ for
categories\footnote[1]{For presentation purposes we shall not deal with size
  issues and assume all the categories are locally small.}. The initial object (when it exists), denoted by $0$,
is the object such that for any other object $A$
there is a unique arrow $0 \xrightarrow{!} A$.  Dually, the terminal object (when it exists),
denoted by $1$, is the object such that for any other object $A$ there is a
unique arrow $A \xrightarrow{!} 1$. As a result of the uniqueness properties, initial and terminal objects are unique up-to
isomorphism.

@@

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$, for all morphisms
$f : X \to Y$. Intuitively, a natural transformation is akin to a polymorphic
function transforming the structure of a functor into the structure of another
functor without assuming what is the type of data contained in them.  For
example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
nothing if the list is empty and the head of the list otherwise can work for all
types $X$ uniformly because it needs not to inspect the data inside the list.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For an endofunctor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}. These are  arrows $f : X \to Y$ in the underlying category $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = a_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
if $F$ has an initial $F$-algebra then this is the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ where
$\InIso : F \mu F \to \mu F$ is the $F$-algebra  witnessing the
isomorphism $F\mu F \iso \mu F$, furthermore  $\InOp : \mu F \to F \mu F$ is the
inverse of $\InIso$. The uniqueness property of initial $F$-algebras states that for any other
$F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra homomorphism, denoted
by $\cata{\alpha_{X}} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
``\emph{catamorphism}'' or ``\emph{fold}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InOp
\end{equation}
As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see.

@@

For example, for a set of states $X$ and a finite set of labels $L$ we can
define a labelled transition system (LTS)~\cite{WinskelM95} on $X$ as a function $X \to BX$
implementing the \emph{transition system} with $BX = L \times X$.  In
particular, for a state $x_{1} \in X$, $c(x_{1})$ returns a pair $(l, x_{2})$
where $l \in L$ is the observable action and $x_{2} \in X$ is the next state.

The category of $B$-coalgebras, denoted $B\Coalg$, is the category where objects
are $B$-coalgebras and morphisms $f : (X, c_{X}) \to (Y, c_{Y})$ are
$B$-coalgebra homomorphisms $f : A \to B$, that is
$ c_{Y} \circ f = F(f) \circ c_{X}$. The terminal object in this category is
called the terminal, or final, $B$-coalgebra. The carrier of this coalgebra
corresponds to the greatest fixed-point for the functor $B$, denoted by
$(\nu B, \OutIso)$ with $\OutIso$ being the final $B$-coalgebra witnessing the
isomorphism and $\OutIso^{\circ} : B\nu B \to \nu B$ being its inverse.

For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
of infinite streams over the set $A$, that is the greatest solution to the equation
$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
a unique $B$-coalgebra homomorphism into the terminal coalgebra
$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
\end{equation}
From the uniqueness property we can derive the fusion law for unfolds:
\begin{equation}
  \label{eq:ana:fusion}
  \ana{c_{Y}} \circ f = \ana{c_{X}}
\end{equation}
for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.

@@

A coalgebra $(C,c)$ is \emph{recursive} if for every algebra $(A, a)$ there is a
\emph{unique} hylo $(C,c ) \rightarrowtail (A, a)$~\cite{CaprettaUV04,AdamekMM19}. We denote these
type of hylos by $\hylo{c}{a}$.

@@

Nevertheless, since \coq{partition} always produces sublists that are strictly
smaller than the input list, it still supports a well-founded recursion scheme,
ensuring the existence of a unique solution. The uniqueness property of the
hylomorphisms yields the following fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
optimisation}~\cite{Wadler90}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. % This, in turn,
% allows us to prove that a recursive hylomorphism is the composition of a
% catamorphism and a recursive anamorphism.

@@

\subsection{Polynomial Functors}
Recursion schemes are formulated using generic functors that encapsulate the
structure of the recursive operator. However, since not all functors have
suitable fixed points, it is necessary to restrict our focus to \emph{strictly
positive functors}.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
%Recursion schemes provide an abstract way to consume and generate data.
We now consider the formalisation of hylomorphisms in Coq. We first formalise
container functors as a tool to represent polynomial functors
(Section~\ref{sec:containers}). Then we formalise algebras for container
extensions (Section~\ref{sec:coq-algebras}). In
Section~\ref{sec:coq-coalg} we formalise coalgebras and put together these notions to
formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:coq-rec-coalgebras}).

@@

Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*...*)
\end{coqcode}
From this definition the universal property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f} that
is an \coq{F}-coalgebra homomorphism then it must be the anamorphism on the same
coalgebra.

@@

\subsubsection{On the subtype of finite elements}
As we mention in
Section~\ref{sec:category-theory}, we define recursive anamorphisms as
hylomorphisms built by using a recursive coalgebra, and the respective
initial \coq{F}-algebra.
In other words, in this development we have defined recursive anamorphisms on
inductive data types.  We might have as well defined them on the subtype of
finite elements of coinductive data types using a predicate which states when an
element of a coinductive data type is finite:
\begin{coqcode}
Definition FinF : GFix F -> Prop := RecF g_out.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic to its corresponding inductive data type \coq{LFix F}.
We show this by
defining a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as hylomorphisms on the
 recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
\begin{coqcode}
Definition rana c := hylo l_in c.
\end{coqcode}
The main advantage of this definition is that recursive anamorphisms compose
with catamorphisms without the need to reason about termination and finiteness
of values.
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}

@@

\subsubsection{Fusing a divide-and-conquer computation}
We now show how we can use program calculation techniques to fuse a traversal
with a  divide-and-conquer algorithm to obtain a new program that only performs
recursion once instead of twice. In particular, we can map a function to the
result of quicksort and obtain a new program which orders the elements and
computes the function on every element by traversing the list only once. The
composition of these two programs is given by the following:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
where \coq{Lmap times_two} is a list map function defined as a hylomorphism, and
\coq{times_two} multiplies every element of the list by two. We can use Coq's
generalised rewriting~\cite{Sozeau09} and \coq{hylo_fusion_l} to fuse \coq{times_two} into
\coq{hylo merge qsplit} which gives the program \coq{hylo (merge \o natural times_two) qsplit}.
In this definition, \coq{natural} defines a natural transformation by applying
\coq{times_two} to the shapes thus multiplying every pivot by two.

@@

In our framework, we formalise $G$ as a container with shape \coq{Sg} and
position \coq{Pg}, then \coq{Memo} is the least fixed-point of the functor
$F X = A \times GX$:
\begin{coqcode}
Instance Memo : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table := LFix Memo.
\end{coqcode}
The operations of the cofree comonad \coq{Memo} are the \coq{Cons}, the \coq{headT} and the \coq{tailT}:
\begin{coqcode}
Definition consT : A * App G Table ~> Table := (* *)
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
Finally, we define a dynamorphism as a hylomorphism from a type \coq{B} into a
type \coq{Table} post-composed with the \coq{headT} map which outputs the last
result from the memory table:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (consT \o pair a id) c.
\end{coqcode}
Note the map \coq{App G Table ~> A} corresponding to a map $G G_{\infty}A \to A$
is not a $G$-algebra.  To recover the $G$-algebra we need to transform it by pairing it with the identity map and
post-composing it with \coq{consT}. This algebra uses this table to lookup
elements, instead of triggering a further recursive call, then elements are
inserted into the memoisation table by the use of \coq{consT} to the result of
applying the algebra. In our example this algebra is called \coq{knapsackA} (see
accompanying code).
% \begin{coqcode}
% Fixpoint memo_knap table wvs :=
%   match wvs with | nil => nil | h :: t =>
%       match lookupT (Nat.pred (fst h)) table with
%       | Some u => (u + snd h)%sint63 :: memo_knapsack table t
%       | None => memo_knapsack table t
%       end
%   end.
% Definition knapsack_alg (wvs : list (nat * int))
%   (x : App NatF (Table NatF int)) : int :=
%   match x with | MkCont sx kx => match sx with
%   | inl tt => fun kx => 0%sint63
%   | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
%   end kx end.
% Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
% \end{coqcode}
Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we
can define knapsack using the recursion scheme for dynamorphisms:
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
The full extracted code coming from this specification is available in the
artefact accompanying this paper, and an interested reader can check that simple
inlining would produce the following:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note that in the extracted code, the recursive calls to \ocaml{f} build the
memoisation table, and that this memoisation table is used to compute the
intermediate results in \ocaml{memo_knap}, which is finally discarded to produce
the final result.

\subsection{Shortcut Deforestation}\label{sec:shortcut}
We now showcase the shortcut deforestation optimisation on lists.  This is used
to remove the creation of intermediate data structures by fusing multiple
recursive traversals into one.  For example, we can use it to show the following
code from Takano and
Meijer~\cite{TakanoM95}
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
which uses that three different hylomorphisms can be fused into one.
Here \coq{sf1} is the composition of \coq{length},
\coq{Lmap f} and \coq{append ys} which are, in particular,
catamorphisms.
